From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-07-28 WOLFHPC15/article/concept.tex
minor fixes

Figure~\ref{fig:ex} gives two examples of stencil computations and introduce the different concepts used in this paper. First, a stencil code is based on a \emph{mesh} which is the discretization of the physical real domain. A mesh is a connected undirected graph without bridges (an edge is a bridge if its removal results in two disconnected graphs), where nodes and edges are linked to form cells (closure). For example, a structured mesh is illustrated in Figure~\ref{fig:ex}, each cell contains four nodes and four edges. Second, \emph{mesh domains} are subset elements of the mesh, such as the center of cells, edges or nodes, for example. In Figure~\ref{fig:ex} two mesh domains are used, the center of the cells (in red) and the edge on the horizontal axe (in blue on the left and right of each cell). A \emph{data} is a quantity to simulate or a temporary data to perform computations. A data is mapped onto a given mesh domain. For example, in the figure, data $A$ and $B$ are mapped onto the center of cells, while $C$ is mapped onto the edges.

@@

The first stencil computation of Figure~\ref{fig:ex} computes $B$ using $A$, while the second one computes $B$ using $C$. Thus, a stencil computation is defined by a set of input data (in those examples only one input data), and a single output data, the result. The \emph{computation domain} is a subset of elements to compute on the output data. In other words a computation domain is a subset of a mesh domain. For example, in the first stencil, the computation of $B$ is performed on plain elements, while dotted elements are not computed. On the other hand, on the second computation, the computation domain of $B$ is exactly equal to the cell mesh domain. The computation domain represents the set of elements on which the computation is performed, in other words, the space loop.

Finally, a stencil computation is composed of a \emph{numerical expression} which indicates how to compute an element of the computation domain (on the output data), using elements of the set of input data. The numerical expression of a stencil computation has the particularity to use not only the element to compute but also a \emph{neighborhood} of this element, also called a \emph{stencil shape}. For example, the stencil shape of the first computation of Figure~\ref{fig:ex} contains direct neighbors on the right, left, top and bottom. Sometimes, the neighborhood shape can also access a different mesh domain, as for example in the second computation example. Actually, in this computation, the neighborhood contains edges on the left and on the right of the cell. As a result, we could imagine for example that the numerical expression of the first example is
\begin{equation*} 
B(x,y) = A(x+1,y)+A(x-1,y)+A(x,y+1)+A(x,y-1),
\end{equation*}
and that the numerical expression of the second example is
\begin{equation*} 
B(x,y) = B(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

@@

%----------------------------------------
\subsection{Multi-stencil programs}
\label{sect:multistencil}
%----------------------------------------
If stencil computations are well defined and a lot studied in the literrature, it is not the case for real case numerical simulations. In this paper, we define what we call a \emph{multi-stencil program}, which actually is a general numerical simulation.

A real case numerical simulation is most of the time not composed of a single stencil computation, but of a set of stencil computations, with one or more stencil shapes (neighboorhood), and onto one or more data. Moreover, a numerical simulation also performs additionnal auxiliary computations which do not involve neighborhood, called local computations. In this paper is targetted this general case of numerical simulations that we call \emph{multi-stencil programs}. The work presented produces a general parallelization structure of the overall simulation, while most stencil solutions produce optimized codes for a single stencil code. As it will be detailed in the related work, this work is complementary to the optimization and parallelization of a single stencil computation.

>>>>

Let introduce some concepts used in this paper. First, a stencil code is based on a \emph{mesh} which is the discretization of the physical real domain. A mesh is a connected undirected graph without bridges (an edge is a bridge if its removal results in two disconnected graphs), where nodes and edges are linked to form cells (closure). For example, a structured mesh is illustrated in Figure~\ref{fig:ex}, each cell contains four nodes and four edges. Second, \emph{mesh domains} are subset elements of the mesh, such as the center of cells, edges or nodes, for example. In Figure~\ref{fig:ex} two mesh domains are used, the center of the cells (in red) and the edge on the horizontal axe (in blue on the left and right of each cell). A \emph{data} is a quantity to simulate or a temporary data to perform computations. A data is mapped onto a given mesh domain. For example, in Figure~\ref{fig:ex}, data $A$ and $B$ are mapped onto the center of cells, while $C$ is mapped onto the edges.

@@

The first stencil computation in Figure~\ref{fig:ex} computes $B$ using $A$, while the second one computes $B$ using $C$. Thus, a stencil computation is defined by a set of input data (in those examples only one input data), and a single output data, the result. The \emph{computation domain} is a subset of elements to compute on the output data. In other words a computation domain is a subset of a mesh domain. For example, in the first stencil, the computation of $B$ is performed on plain elements, while dotted elements are not computed. On the other hand, on the second computation, the computation domain of $B$ is exactly equal to the cell mesh domain. The computation domain represents the set of elements on which the computation is performed, in other words, the space loop.

Finally, a stencil computation is composed of a \emph{numerical expression} which indicates how to compute an element of the computation domain (on the output data), using elements of the set of input data. The numerical expression of a stencil computation has the particularity to use not only the element to compute but also a \emph{neighborhood} of this element, also called a \emph{stencil shape}. For example, the stencil shape of the first computation in Figure~\ref{fig:ex} contains direct neighbors on the right, left, top and bottom. Sometimes, the neighborhood shape can also access a different mesh domain, as for example in the second computation example. Actually, in this computation, the neighborhood contains edges on the left and on the right of the cell. As a result, we could imagine for example that the numerical expression of the first example is
\begin{equation*} 
B(x,y) = A(x+1,y)+A(x-1,y)+A(x,y+1)+A(x,y-1),
\end{equation*}
and that the numerical expression of the second example is
\begin{equation*} 
B(x,y) = B(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

@@

%----------------------------------------
\subsection{Multi-stencil programs}
\label{sect:multistencil}
%----------------------------------------
While stencil computations are well defined and a lot studied in the literrature, it is not the case for real case numerical simulations. In this paper, we define what we call a \emph{multi-stencil program}, which actually is a general numerical simulation.

A real case numerical simulation is most of the time not composed of a single stencil computation, but of a set of stencil computations, with one or more stencil shapes (neighboorhood), and onto one or more data. Moreover, a numerical simulation also performs additionnal auxiliary computations which do not involve neighborhood, called local computations. This paper targets this general case of numerical simulations that we call \emph{multi-stencil programs}. The work presented produces a general parallelization structure of the overall simulation, while most stencil solutions produce optimized codes for a single stencil code. As it will be detailed in the related work, this work is complementary to the optimization and parallelization of a single stencil computation.

### 2015-07-28 WOLFHPC15/article/ms_msc.tex
minor fixes

\paragraph{Computations description} The most important part of MSL is the description of the simulation computations. The language follows the three definitions given in Section~\ref{sect:multistencil}, however as MSL is not made to express numerical computations, the element $exp$ of definitions~\ref{eq:st} and~\ref{eq:loc} is not given in the language. Moreover, to propose a syntax close to an imperative code, information are given in a different order.
Thus, the description of a computation consists of the identifier of the data to compute, the computation domain (which is a subpart of the mesh domain of the data) on which it is computed into the brackets, an identification name of the computation, and a set of identifiers for the data read by the computation. For each data read by the computation brackets indicate the stencil shape identifier used on the data. If the computation is local brackets stay empty. In the example below, the data $b$ is computed on the computation domain $allcell$ by the computation $c_0$ which read data $a$ without neighborhood shape.
\begin{lstlisting}[basicstyle=\small,mathescape,frame=single,language=C++]
computations:
  b[allcell]=c0(a[])
  c[alledgex]=c1(b[n1])
  d[alledgex]=c2(c[])
  e[alledgey]=c3(c[])
  f[allcell]=c4(d[n1])
  g[alledgey]=c5(e)
  h[alledgex]=c6(f)
  i[allcell]=c7(g[],h[])
  j[partedgex]=c8(i[n1])
\end{lstlisting}

@@

\paragraph{Creation of the synchronized ordered list of computations} COnsidering a data parallelization of the simulation, from the ordered list of computations $\Gamma$ can be detect needed synchronizations. A synchronization is needed each time a data read by a stencil computation has been written by a previous computation and if the computation domains intersect. Actually, as a stencil computation needs neighborhood values, values computed by different processors or cores are needed.

>>>>

\paragraph{Computations description} The most important part of MSL is the description of the simulation computations. The language follows the three definitions given in Section~\ref{sect:multistencil}. However as MSL is not made to express numerical computations, the term $exp$ of Definition~\ref{eq:st} and Definition~\ref{eq:loc} is not given in the language. Moreover, to propose a syntax close to an imperative code, information are given in a different order.
Thus, the description of a computation consists of the identifier of the data to compute, the computation domain (which is a subpart of the mesh domain of the data) on which it is computed into the brackets, an identification name of the computation, and a set of identifiers for the data read by the computation. For each data read by the computation brackets indicate the stencil shape identifier used on the data. If the computation is local brackets stay empty. In the example below, the data $b$ is computed on the computation domain $allcell$ by the computation $c_0$ which read data $a$ without neighborhood shape.
\begin{lstlisting}[basicstyle=\small,mathescape,frame=single,language=C++]
computations:
  b[allcell]=c0(a[])
  c[alledgex]=c1(b[n1])
  d[alledgex]=c2(c[])
  e[alledgey]=c3(c[])
  f[allcell]=c4(d[n1])
  g[alledgey]=c5(e)
  h[alledgex]=c6(f)
  i[allcell]=c7(g[],h[])
  j[partedgex]=c8(i[n1])
\end{lstlisting}

@@

\paragraph{Creation of the synchronized ordered list of computations} Considering a data parallelization of the simulation, from the ordered list of computations $\Gamma$ can be detect needed synchronizations. A synchronization is needed each time a data read by a stencil computation has been written by a previous computation and if the computation domains intersect. Actually, as a stencil computation needs neighborhood values, values computed by different processors or cores are needed.
